\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}

### 2021-08-05 move_schedules.tex
Minor changes to fibered barcode section

The dimension function is simply the the function which maps every point $a \in \mathbb{R}^2$ to $\mathrm{dim}(M_a)$. It is a simple and easy to visualize invariant, but is unstable and yields no information about the persistent features of $M$. 
The multigraded Betti numbers constitute a natural and important class of invariants often studied in algebraic geometry commutative algebra, though both their theory and computation is beyond the scope of this effort---the interested reader is referred to~\cite{lesnick2015interactive, carlsson2009theory} for more details. For our purposes, the the $i^{\text{th}}$-graded Betti number of $M$ is simply a function $\beta_i(M): \mathbb{R}^n \to \mathbb{N}$. The fibered barcode, denoted as $\mathcal{B}(M)$, is defined as follows: 
\begin{definition}[Fibered barcode]
	%\in \overline{\mathcal{L}}
	The fibered barcode $\mathcal{B}(M)$ of a bipersistence module $M$ is the map which sends each line $L$ with non-negative slope to the barcode $\mathcal{B}_L(M)$: 
%	$$ L \mapsto \mathcal{B}_L(M) $$
$$ \mathcal{B}(M) = \{ \; B_L(M) : L \in \mathbb{R} \times \mathbb{R}^{+} \; \}$$
\end{definition} 
\noindent Thus, given a bipersistence module $M$, the fibered barcode of $M$ is the 2-parameter family of barcodes given by restricting $M$ to the of set affine lines with non-negative slope in $\mathbb{R}^2$.

Although the fiber barcode is a very intuitive invariant, it is not clear how one might go about computing it efficiently. One could compute the 1-D persistence associated to some fixed linear combination of two given filter functions; each linear combination parameterizes a line $L \in \mathbb{R} \times \mathbb{R}^{+}$ which $M$ restricts too. However, this an $O(m^3)$ time computation, which is prohibitive for exploratory data analysis purposes. In what follows, we describe the reparameterization introduced by~\cite{lesnick2015interactive} that fully characterizes $\mathcal{B}(M)$.

@@

Let $S = \mathrm{supp}\,\beta_0(M) \, \cup \, \mathrm{supp}\,\beta_1(M)$, where the functions $\beta_0(M), \beta_1(M)$ are $0^{\text{th}}$ and $1^{\text{st}}$ bigraded Betti numbers of $M$, respectively. 
The main mathematical result from~\cite{lesnick2015interactive} is a characterization of the barcodes $\mathcal{B}_L(M)$, for any $L \in \mathcal{L}$, in terms of a set of \emph{barcode templates} $\mathcal{T}$ computed at every 2-cell in $\mathcal{A}(M)$.
 More formally, for any line $L \in \overline{\mathcal{L}}$ and $e$ any 2-cell in $\mathcal{A}(M)$ whose closure contains the dual of $L$ under point-line duality, the 1-parameter restriction of the persistence module $M$ induced by $L$ is given by: 
 \begin{equation}
 \mathcal{B}_L(M) = \{ [\,\mathrm{push}_L(a), \, \mathrm{push}_L(b)\,) \mid (a,b) \in \mathcal{T}^e, \mathrm{push}_L(a) < \mathrm{push}_L(b)  \} 
\end{equation}
Minor additional conditions are needed for handling completely horizontal and vertical lines. 
The importance of this theorem for this effort lies in the fact that the fibered barcodes are completely defined from the precomputed barcode templates $\mathcal{T}$---once every barcode template $\mathcal{T}^e$ has been computed and augmented onto $\mathcal{A}(M)$, $\mathcal{B}(M)$ is completely characterized, and the barcodes $\mathcal{B}_L(M)$ associated to a 1-D filtration induced by \emph{any} choice of $L$ can be efficiently computed via a point-location query on $\mathcal{A}(M)$ and a $O(\lvert \mathcal{B}_L(M) \rvert)$ application of the push map.

@@

% Indeed, it was shown in~\cite{lesnick2015interactive} that the fibered barcodes $$ of $M$ (i.e. the collection of barcodes of 1D affine slices of $M$) and the rank invariant determine each other.     
\subsubsection*{Computing the fibered barcode}
%Lesnick and Wright introduced the Rank Invariant Visualization and Exploration Tool (RIVET)~\cite{lesnick2015interactive}, which is both a tool and a collection of computational and mathematical theory which enables exploratory analysis of 2D persistence modules built from arbitrary bifiltrations. From a computational perspective, the core novelty of the RIVET approach is its ability to compute the fibered barcodes efficiently from precomputed information. 
The three dominant subcomputations needed to compute the fibered barcode using the reparameratization discussed above are the following: 
\begin{enumerate}
	\item Computing the bigraded Betti numbers $\beta(M)$ of a given bipersistence module $M$ 
	\item Constructing a line arrangement $\mathcal{A}(M)$ induced by critical lines derived from (1) 
	\item Augmenting $\mathcal{A}(M)$ with a \emph{barcode template} $\mathcal{T}^e$ at every 2-cell $e \in \mathcal{A}(M)$
	%, yielding the \emph{augmented arrangement} $\mathcal{A}^{\sbullet}(M)$ 
\end{enumerate}
Computing (1) takes approximately $\approx O(m^3)$ using a matrix algorithm similar to Algorithm~\ref{alg:reduce}~\cite{lesnick2019computing}. Constructing and storing the line arrangement $\mathcal{A}(M)$ with $n$ lines and $k$ vertices is related to the \emph{line segment intersection problem}, which known algorithms in computational geometry can solve in (optimal) output-sensitive $O((n+k) \log n)$ time (see~\cite{boissonnat2000robust} for an overview). In terms of memory complexity, the number of 2-cells in $\mathcal{A}(M)$ is upper bounded by $O(\kappa^2)$, where $\kappa$ is a coarseness parameter associated with the computation of $\beta(M)$. 
By combining this bound with a bound on the number of transpositions needed to traverse a particular sequence of 2-cells in $\mathcal{A}(M)$,~\cite{lesnick2015interactive} show that the computation of the barcodes templates $\mathcal{T}$ from step (3) using vineyard updates requires on the order of $O(m^3 \kappa + m \kappa^2 \log \kappa)$ elementary operations and $O(m \kappa^2)$ storage.

Of the three steps above, the barcode template computation is both the highest complexity and most demanding computation in practice. The number of 2-cells in $\mathcal{A}(M)$ is on the order $O(\kappa^2)$, and $\kappa$ itself is on the order of $O(m^2)$ in the worst case. 
This suggests the computation of the barcode templates is on the order $O(m^5)$, which is effectively intractable for moderately-sized filtrations. In practice, the external stability result from~\cite{landi2014rank} justifies the use of a simple procedure which approximates the module $M$ with a smaller module $M'$ using a grid-like coarsening procedure. Two binning parameters associated with the coarsening procedure, $x_{bin}$ and $y_{bin}$, together determine the degree of the approximation. The coarsening procedure allows the practitioner to restrict the size of $\kappa$ to a relatively small constant, which in-turn dramatically reduces the size of $\mathcal{A}(M)$. 
With this in mind, given a multifiltration with $m$ simplices, the entire computation for approximating the fibered barcode invariant reduces to an $\approx O(m^3)$ algorithm.

There are several approaches one can use to compute $\mathcal{T}$, the simplest being to run Algorithm~\ref{alg:reduce} independently on the 1-D filtration induced by the duals of some set of points (e.g. the barycenters) lying in the interior of the 2-cells of $\mathcal{A}(M)$.  
The approach taken by~\cite{lesnick2015interactive} is to use the $R = DV$ decomposition computed at some adjacent 2-cell $e \in \mathcal{A}(M)$ to speed up the computation of an adjacent cell $e' \in \mathcal{A}(M)$. More explicitly, define the \emph{dual graph} of $\mathcal{A}(M)$ to be the undirected graph $G$ which has a vertex for every 2-cell $e \in \mathcal{A}(M)$ and an edge for each adjacent pair of cells $e, e' \in \mathcal{A}(M)$.
Each vertex in $G$ is associated with a barcode template $\mathcal{T}^e$, and the computation of $\mathcal{T}$ now reduces to computing a path $\Gamma$ on $G$ which visits each vertex at least once. To minimize the computation time, assume the $n$ edges of $G$ are endowed with non-negative weights $W = w_1, w_2, \dots, w_n$ whose values $w_i \in \mathbb{R}_{+}$ represent some notion of distance. The optimal path $\Gamma^\ast$ is the one of minimal length with respect to $W$ which visits every vertex of $G$ at least once. There is a known $\frac{3}{2}$-approximation that can be computed efficiently which reduces the problem to the traveling salesman problem on a metric graph~\cite{christofides1976worst}. In the next paragraph, we include a small example to illustrate many of the concepts and definitions discussed so far below. 
\\
\\
\textbf{Noisy circle example}
Consider a small set of points distributed noisily around $S^1$ that contains a few strong outliers, shown on the left side of Figure~\ref{fig:bifiltration_ex2}. Filtering this data set with respect to the Rips parameter and the complement of a kernel density estimate yields a bifiltration whose dimension function and bigraded Betti numbers are shown in the middle figure. The gray areas indicate the region where the dimension function is constant. In particular, the light gray area of the middle plot shows where $\mathrm{dim}_1(M) = 1$, indicating where in the multifiltration a persistent 1-cycle was detected.
On the right side, the space dual to the space in the middle is shown. The black lines represent the lines of $\mathcal{A}(M)$, the blue dashed-lines the edges of the dual graph $\mathcal{G}$, and the barycenters of the 2-cells in $\mathcal{A}(M)$ depicting where the barcodes templates $\mathcal{T}^e$ are computed are drawn as orange points. 
\begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{bifiltration_ex2}
	\caption{Bipersistence example on an $8 \times 8$ coarsened grid. On left, the input data, colored by density. In the middle, the bigraded Betti numbers $\beta_0(M)$ and $\beta_1(M)$ (green and red, respectively), the Hilbert function (gray), and a line $L$ emphasizing the persistence of features with high density. On the right, the line arrangement $\mathcal{A}(M)$ lying in the dual space $D_p(\alpha)$ derived from the $\beta(M)$. }
	\label{fig:bifiltration_ex2}
\end{figure}
The shortest path $\Gamma^\ast$ spanning the vertices of $G$ is shown by the solid rainbow colored lines, with the traversal beginning in the top-left cell (at the red edge) and ending at the bottom-right cell (the indigo edge). Traversing $\Gamma^\ast$ can be done via a simple depth-first search, though note that a $R=DV$ decomposition needs to be stored at each degree-3 or higher vertex in $\Gamma^\ast$ to avoid redundant computations as the traversal recurses. 
\\

>>>>

The dimension function is simply the the function which maps every point $a \in \mathbb{R}^2$ to $\mathrm{dim}(M_a)$. It is a simple and easy to visualize invariant, but is unstable and yields no information about the persistent features of $M$. 
The multigraded Betti numbers constitute a natural and important class of invariants often studied in algebraic geometry commutative algebra, though both their theory and computation is beyond the scope of this effort---the interested reader is referred to~\cite{lesnick2015interactive, carlsson2009theory} for more details. For our purposes, the $i^{\text{th}}$-graded Betti number of $M$ is simply a function $\beta_i(M): \mathbb{R}^n \to \mathbb{N}$ whose values $\beta_i^z(M)$ indicate the the number of elements at grade $z$ in a basis of the $i^{\text{th}}$ module in a free resolution for $M$. The fibered barcode, denoted as $\mathcal{B}(M)$, is defined as follows: 
\begin{definition}[Fibered barcode]
	%\in \overline{\mathcal{L}}
	The fibered barcode $\mathcal{B}(M)$ of a bipersistence module $M$ is the map which sends each line $L$ with non-negative slope to the barcode $\mathcal{B}_L(M)$: 
%	$$ L \mapsto \mathcal{B}_L(M) $$
$$ \mathcal{B}(M) = \{ \; B_L(M) : L \in \mathbb{R} \times \mathbb{R}^{+} \; \}$$
\end{definition} 
\noindent Thus, given a bipersistence module $M$, the fibered barcode of $M$ is the 2-parameter family of barcodes given by restricting $M$ to the of set affine lines with non-negative slope in $\mathbb{R}^2$.

Although the fiber barcode is a very intuitive invariant, it is not clear how one might go about computing it efficiently. One could compute the 1-D persistence associated to some fixed linear combination of two given filter functions; each linear combination parameterizes a line $L \in \mathbb{R} \times \mathbb{R}^{+}$ which $M$ restricts too. However, this an $O(m^3)$ time computation, which is prohibitive for exploratory data analysis purposes. Ideally, one would want to continuously vary $L \in \mathbb{R} \times \mathbb{R}^{+}$ interactively, inspecting the corresponding changes to the barcodes immediately. To enable this, ~\cite{lesnick2015interactive} introduced an elegant reparameterization that not only fully characterizes $\mathcal{B}(M)$, but also yields an efficient method for rendering $\mathcal{B}_L(M)$ in real time. In the next section, we briefly summarize this reparameterization.

@@

Let $S = \mathrm{supp}\,\beta_0(M) \, \cup \, \mathrm{supp}\,\beta_1(M)$, where the functions $\beta_0(M), \beta_1(M)$ are $0^{\text{th}}$ and $1^{\text{st}}$ bigraded Betti numbers of $M$, respectively. 
The main mathematical result from~\cite{lesnick2015interactive} is a characterization of the barcodes $\mathcal{B}_L(M)$, for any $L \in \mathcal{L}$, in terms of a set of \emph{barcode templates} $\mathcal{T}$ computed at every 2-cell in $\mathcal{A}(M)$.
 More formally, for any line $L \in \overline{\mathcal{L}}$ and $e$ any 2-cell in $\mathcal{A}(M)$ whose closure contains the dual of $L$ under point-line duality, the 1-parameter restriction of the persistence module $M$ induced by $L$ is given by: 
 \begin{equation}
 \mathcal{B}_L(M) = \{ [\,\mathrm{push}_L(a), \, \mathrm{push}_L(b)\,) \mid (a,b) \in \mathcal{T}^e, \mathrm{push}_L(a) < \mathrm{push}_L(b)  \} 
\end{equation}
Minor additional conditions are needed for handling completely horizontal and vertical lines. 
The importance of this theorem lies in the fact that the fibered barcodes are completely defined from the precomputed barcode templates $\mathcal{T}$---once every barcode template $\mathcal{T}^e$ has been computed and augmented onto $\mathcal{A}(M)$, $\mathcal{B}(M)$ is completely characterized, and the barcodes $\mathcal{B}_L(M)$ associated to a 1-D filtration induced by \emph{any} choice of $L$ can be efficiently computed via a point-location query on $\mathcal{A}(M)$ and a $O(\lvert \mathcal{B}_L(M) \rvert)$ application of the push map.

@@

% Indeed, it was shown in~\cite{lesnick2015interactive} that the fibered barcodes $$ of $M$ (i.e. the collection of barcodes of 1D affine slices of $M$) and the rank invariant determine each other.     
\subsubsection*{Computing the fibered barcode}
%Lesnick and Wright introduced the Rank Invariant Visualization and Exploration Tool (RIVET)~\cite{lesnick2015interactive}, which is both a tool and a collection of computational and mathematical theory which enables exploratory analysis of 2D persistence modules built from arbitrary bifiltrations. From a computational perspective, the core novelty of the RIVET approach is its ability to compute the fibered barcodes efficiently from precomputed information. 
The three dominant subcomputations needed to compute the fibered barcode using the reparameratization discussed above are the following: 
\begin{enumerate}
	\item Computing the bigraded Betti numbers $\beta(M)$ of a given bipersistence module $M$ 
	\item Constructing a line arrangement $\mathcal{A}(M)$ induced by critical lines derived from (1) 
	\item Augmenting $\mathcal{A}(M)$ with a \emph{barcode template} $\mathcal{T}^e$ at every 2-cell $e \in \mathcal{A}(M)$
	%, yielding the \emph{augmented arrangement} $\mathcal{A}^{\sbullet}(M)$ 
\end{enumerate}
Computing (1) takes approximately $\approx O(m^3)$ using a matrix algorithm similar to Algorithm~\ref{alg:reduce}~\cite{lesnick2019computing}. Constructing and storing the line arrangement $\mathcal{A}(M)$ with $n$ lines and $k$ vertices is related to the \emph{line segment intersection problem}, which known algorithms in computational geometry can solve in (optimal) output-sensitive $O((n+k) \log n)$ time (see~\cite{boissonnat2000robust} for an overview). In terms of memory complexity, the number of 2-cells in $\mathcal{A}(M)$ is upper bounded by $O(\kappa^2)$, where $\kappa$ is a coarseness parameter associated with the computation of $\beta(M)$. 
By combining this bound with a bound on the number of transpositions needed to traverse a particular sequence of 2-cells in $\mathcal{A}(M)$, the analysis from~\cite{lesnick2015interactive} shows that the computation of the barcodes templates $\mathcal{T}$ from step (3) using vineyard updates requires on the order of $O(m^3 \kappa + m \kappa^2 \log \kappa)$ elementary operations and $O(m \kappa^2)$ storage.

Of the three steps above, the barcode template computation is both the highest complexity and most demanding computation in practice. The number of 2-cells in $\mathcal{A}(M)$ is on the order $O(\kappa^2)$, and $\kappa$ itself is on the order of $O(m^2)$ in the worst case. 
This suggests the computation of the barcode templates is on the order $O(m^5)$, which is effectively intractable for even moderately-sized filtrations. In practice, the external stability result from~\cite{landi2014rank} justifies the use of a simple procedure which approximates the module $M$ with a smaller module $M'$ using a grid-like coarsening procedure. Two binning parameters associated with the coarsening procedure, $x_{bin}$ and $y_{bin}$, together determine the degree of the approximation. The coarsening procedure allows the practitioner to restrict the size of $\kappa$ to a relatively small constant, which in-turn dramatically reduces the size of $\mathcal{A}(M)$.

There are several approaches one can use to compute $\mathcal{T}$, the simplest being to run Algorithm~\ref{alg:reduce} independently on the 1-D filtration induced by the duals of some set of points (e.g. the barycenters) lying in the interior of the 2-cells of $\mathcal{A}(M)$.  
The approach taken by~\cite{lesnick2015interactive} is to use the $R = DV$ decomposition computed at some adjacent 2-cell $e \in \mathcal{A}(M)$ to speed up the computation of an adjacent cell $e' \in \mathcal{A}(M)$. More explicitly, define the \emph{dual graph} of $\mathcal{A}(M)$ to be the undirected graph $G$ which has a vertex for every 2-cell $e \in \mathcal{A}(M)$ and an edge for each adjacent pair of cells $e, e' \in \mathcal{A}(M)$.
Each vertex in $G$ is associated with a barcode template $\mathcal{T}^e$, and the computation of $\mathcal{T}$ now reduces to computing a path $\Gamma$ on $G$ which visits each vertex at least once. To minimize the computation time, assume the $n$ edges of $G$ are endowed with non-negative weights $W = w_1, w_2, \dots, w_n$ whose values $w_i \in \mathbb{R}_{+}$ represent some notion of distance. The optimal path $\Gamma^\ast$ is the one of minimal length with respect to $W$ which visits every vertex of $G$ at least once. There is a known $\frac{3}{2}$-approximation that can be computed efficiently which reduces the problem to the traveling salesman problem on a metric graph~\cite{christofides1976worst}.
\\
\\
\textbf{Noisy circle example} We include a small example to illustrate many of the concepts and definitions discussed so far. 
Consider a small set of points distributed noisily around $S^1$ that contains a few strong outliers, shown on the left side of Figure~\ref{fig:bifiltration_ex2}. Filtering this data set with respect to the Rips parameter and the complement of a kernel density estimate yields a bifiltration whose dimension function and bigraded Betti numbers are shown in the middle figure. The gray areas indicate the region where the dimension function is constant; the lighter gray area where $\mathrm{dim}_1(M) = 1$, indicating where a persistent loop was detected.
On the right side, the space dual to the space in the middle is shown. The black lines represent the lines of $\mathcal{A}(M)$, the blue dashed-lines the edges of the dual graph $\mathcal{G}$, and the orange points the barycenters of the 2-cells in $\mathcal{A}(M)$, depicting where the barcodes templates $\mathcal{T}^e$ are parameterized. 
\begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{bifiltration_ex2}
	\caption{Bipersistence example on an $8 \times 8$ coarsened grid. On left, the input data, colored by density. In the middle, the bigraded Betti numbers $\beta_0(M)$ and $\beta_1(M)$ (green and red, respectively), the Hilbert function (gray), and a line $L$ emphasizing the persistence of features with high density. On the right, the line arrangement $\mathcal{A}(M)$ lying in the dual space $D_p(\alpha)$ derived from the $\beta(M)$. }
	\label{fig:bifiltration_ex2}
\end{figure}
The shortest path $\Gamma^\ast$ spanning the vertices of $G$ is shown by the solid rainbow colored lines, with the traversal beginning in the top-left cell (at the red edge) and ending at the bottom-right cell (the indigo edge). Traversing $\Gamma^\ast$ can be done via a simple depth-first search, though note that a $R=DV$ decomposition needs to be stored at each degree-3 or higher vertex in $\Gamma^\ast$ to avoid redundant computations as the traversal recurses. 
\\
